---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r include = FALSE}
knitr::opts_chunk$set(warning=FALSE,message=FALSE)

```

For this exercise, we shall use the following libraries to load, process and produce the results for each of the section below:

* **lubridate** : For managing date variables.
* **readr**     : For reading in data into a data frame from source.
* **dplyr**     : For data manipulation.
* **purrr**     : For functional programming capabilities.
* **ggplot2**   : For data visualization.
* **skimr**     : For summary statistics of the data set.
* **kableExtra**: For displaying tables in HTML documents.

Let's install and load these packages into the R Environment:

```{r load_packages}
pkgs <- c('lubridate', 'readr', 'dplyr', 'ggplot2', 'purrr', 'skimr', 'kableExtra')
install.packages(pkgs[!(pkgs %in% installed.packages())]) # Install any missing packages
purrr::walk(pkgs, function(pkg) require(pkg, character.only = TRUE)) # Load packages

```

## Loading and preprocessing the data

Let's load the data from source and structure it into a format that we can use.

```{r data_processing_part1}
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
data_name <- 'activity.zip'

if(!file.exists(data_name)) { # Download data if required
  download.file(url = url, destfile = data_name)
}

activity <- read_csv(data_name) # read_csv can read and load zipped contents.

# Analyse the data for anomolies.
(skim_with(numeric = sfl(hist = NULL)))(activity)

number_of_NA <- sum(length(is.na(activity$steps)))

```

From the analysis, we can see that the *steps* variable has ```r format(round(((is.na(activity$steps) %>% sum) / nrow(activity) * 100), 2), nsmall = 2)```% as *NA's*. By looking at the Mean and Median values, we can also see that its heavily skewed to the right. 

At this stage, we shall keep the NA values as-is indicating that the use of the average value may not be the best use of measure for the NA values. We shall handle the NA values under the *Imputing Missing Values* section.


With the pre-processing done, our data for analysis looks as follows:

```{r data_processing_summary} 

activity %>% 
  kbl() %>% 
  kable_classic(
    bootstrap_options = "striped", font_size = 10) %>% 
  scroll_box(height = "200px")

```

## What is mean total number of steps taken per day?

For this part of the assignment, I have used the cleaned data that replaced the NA values of the steps variable with the median.

Answering this question will take a three-step approach:

**1. Calculate the total number of steps taken per day**

```{r total_daily_steps}

total_daily_steps <- activity %>% 
  group_by(date) %>% 
  summarize(total_steps = sum(steps, na.rm = TRUE)) %>% 
  ungroup()

total_daily_steps %>%
  kbl() %>% kable_classic(
    bootstrap_options = "striped", font_size = 10) %>%
  scroll_box(height = "200px")


```

**2. Make a histogram of the total number of steps taken per day**

Using our total daily steps dataset, we can plot a histogram of the results.

```{r histogram_total_steps}
total_daily_steps %>% 
  ggplot(aes(x = total_steps)) + 
  geom_histogram() + 
  labs(
    x = 'Total Steps', 
    y = 'Frequency',
    title = 'Histogram on the total steps per day',
    caption = 'Period of October and November 2021'
  ) + 
  theme_bw()

```

**3. Calculate and report the mean and median of the total number of steps taken per day**

For the period between October and November 2021, the statistics of interest are as below:

* Median number of steps : `r format(round(median(activity$steps, na.rm = TRUE), 2), nsmall = 2)` steps
* Mean number of steps   : `r format(round(mean(activity$steps, na.rm = TRUE), 2), nsmall = 2)` steps.

## What is the average daily activity pattern?

For this exercise, we shall:

**1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

```{r time_series}

# Summary of step count by 5-minute interval
time_series_data <- activity %>% 
  group_by(interval) %>% 
  summarize(average_steps = mean(steps, na.rm = TRUE)) %>% 
  ungroup()

# Interval with highest step count
max_steps_interval <- time_series_data %>%
  filter(average_steps == max(time_series_data$average_steps, na.rm = TRUE))

# Plotting of data
time_series_data %>% ggplot(aes(x = interval, y = average_steps)) + 
  geom_point() + 
  geom_line(col = 'blue') + 
  labs(
    x = 'Interval', 
    y = 'Average Number of Steps', 
    title = 'Average number of steps taken by 5-minute interval',
    caption = 'Period of October and November 2021'
  ) + 
  geom_vline(xintercept = max_steps_interval$interval, 
             col = 'red', 
             linetype = 'dotted') + 
  geom_text(
    aes(
      x = max_steps_interval$interval, 
      label=paste("Interval", max_steps_interval$interval, ", av. steps ", round(max_steps_interval$average_steps, 2)),
                  y=10), 
    colour="red", angle=0, size=4, hjust = -0.01) + 
  theme_bw()

```

**2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

From the above graph, its clear to see that the maximum number of average steps was **`r round(max_steps_interval$average_steps,2)`**, at the interval **`r max_steps_interval$interval`**.

## Imputing missing values

**1. Calculate and report on the total number of missing values in the dataset.**

In the pre-processing stage, we estimated ```r format(round(((is.na(activity$steps) %>% sum) / nrow(activity) * 100), 2), nsmall = 2)```% of the number of steps constituted *NA's*, and that the distribution was right skewed.

```{r number_of_NA}
activity %>%
  summarize(Number_of_NAs = sum(is.na(activity$steps)), 
            '% as NAs' = Number_of_NAs / nrow(activity) * 100)

```

**2. Devise a strategy for filling in the missing values in the dataset.**

To keep it simple for this exercise, we shall replace the NA's with the **Median** number of steps across the dataset.

**3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**

We shall duplicate the original activity dataset and impute the missing values with the strategy mentioned in 2. above, and display a sample of the result below.

```{r data_processing_part2}
new_activity <- activity
median_steps <- median(activity$steps, na.rm = TRUE)

new_activity$steps <- replace(
  x = activity$steps, 
  list = which(is.na(activity$steps), arr.ind = TRUE), 
  values = median_steps)

head(new_activity)

```

**4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.**

```{r new_activity_histogram}

new_activity %>% 
  group_by(date) %>% 
  summarize(total_steps = sum(steps)) %>% 
  ggplot(aes(x = total_steps)) + 
  geom_histogram() + 
  labs(
    x = 'Total Steps', 
    y = 'Frequency',
    title = 'Histogram on the total steps per day',
    caption = 'Period of October and November 2021'
  ) + 
  theme_bw()

```

Given that we are using the median number of steps (`r median_steps` steps), there does not seem to be much difference in the nature of the histogram compared to the initial that was drawn with the original data set.

We can also compare the means and median values between the original dataset and this new one.

```{r compare_means_medians}
original <- activity %>% summarize(mean = mean(steps, na.rm = TRUE), 
                                   median = median(steps, na.rm = TRUE), 
                                   source = 'original')

new <- new_activity %>% summarize(mean = mean(steps), 
                                  median = median(steps), 
                                  source = 'new')
print(bind_rows(original,new))

```
Between the original dataset (that had NAs) and the new dataset (with imputed median values), we can see that the median values have no variation. However, there is a slight variation between the mean values.

## Are there differences in activity patterns between weekdays and weekends?
